---
title: "E-cigarette use in Great Britain between 2014 and 2023"
output: html_document
date: "2024-11-21"

install.packages("prettydoc")
library(prettydoc)

output: 
  prettydoc::html_pretty:
    theme: leonids
    toc: yes
    toc_float:
      collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Data Origins
This project is focused on visualising the change in e-cigarette use in Great Britain between 2014-2023. The data was found on the Office for National Statistics website, where each year the smoking habits of Great Britain are quantified in using two surveys, and a statistical bulletin is published. This data in particular has been collected over 10 years and collated by the Office for National Statistics in a statistical bulletin published on 01/10/2024.

# Research Question
### How has e-cigarette use changed since 2014?

# Importing the Data
To upload the dataset into R, I used the 'here' function as this allows the code to work on any computer which has downloaded the 'PSY6422_Project' folder. 
As the data is an Excel workbook with a lot of sheets, I only wanted to upload and analyse a particular sheet. The one I chose was 'Table_1' as this looks at e-cigarette use in the UK by sex and age. 
```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(repos = c(CRAN = "https://cran.rstudio.com/"))
```

```{r Load the data, warning=FALSE}
# need to load the dataset into R

# first we need to install and load the required packages
install.packages("here")
install.packages("tidyverse")
library(here)
library(tidyverse)

# the raw dataset is an excel workbook with many sheets so for this project we're specifying one sheet to upload into R
# so need to install packages to load one sheet from the raw dataset
install.packages("readxl") 
install.packages("openxlsx")
library(readxl)
library(openxlsx)

# need to tell R the file path to choose
file_path <- here("Data", "ecigaretteuseingreatbritain.xlsx")

# import data using the given file path
# as the raw dataset is so large, we're specifying which sheet in the raw dataset to use
# we're also going to skip the first 11 rows of the sheet as this is just information about the table
data <- read_xlsx(file_path, sheet = "Table_1", skip = 11)

# view the imported data
head(data)

```