---
title: "Number of convicted homicide suspects, by sex and age, between 1969 and 2023"
author: 'Registration number: 240188795'
date: "2024-11-21"
output:
  html_document:
    theme: cosmo
    toc: yes
    toc_float: 
      collapsed: true
---

```{r setup, include=FALSE, messages=FALSE, results=FALSE}
knitr::opts_chunk$set(echo = TRUE) #to avoid error message: setting CRAN without a mirror
options(repos = c(CRAN = "https://cran.rstudio.com/")) 
#repos specifies the repository 
# this URL is the designated URL for CRAN 
```

# Data Origins
This project is focused on visualising the number of suspects convicted of homocide by the types of homocide between January 1969 and March 2023, categorised by sex and age. This visualisation allows us to compare male and female homocide convictions and see which age (in both sexes) the most homocides are convicted at.

# Research Question
### Are males or females historically more likley to commit homicide?
This visualisation aims to show a clear distinction in the different types of homicide and their trends over the years.

# Importing the Data
To upload the dataset into R, I used the 'here' function as this allows the code to work on any computer which has downloaded the 'Data' folder for this project. 
As the data is an Open Document Spreadsheet with a lot of sheets, I only wanted to upload and analyse a particular sheet. The one I chose was 'Table_26' as this includes suspects for all offences initially ruled a homicide between January 1969 and March 2023. 

```{r Load the data, warning=FALSE, message = FALSE, results=FALSE}
# First, we need to install and load the necessary packages

install.packages("here")
install.packages("tidyverse")
install.packages("dplyr")
install.packages("readxl") 
library(here)
library(tidyverse)
library(readxl)
library(dplyr)

# Now we need to tell R the file path to choose
file_path <- here("Data", "homicideyemarch23appendixtablesfinal1.xlsx")

# As the raw dataset is so large, we're specifying which sheet to use
# We're also going to skip the first 6 rows of the sheet as this is just information about the table

data <- read_excel(file_path, sheet = "Worksheet 31", skip = 7)

# View the imported data
head(data)
```

# Cleaning the Data
In the raw data, this worksheet shows the number of suspects convicted on different counts of homicide, but also the percentages of suspects convicted of homicide by age and sex. We do not want to include the percentage data as this is not representative unless we know the raw numbers the percentages are generated from (which we do not) so we want to remove this. We also want to remove any other unnecessary information in the dataset. For example, there are a few rows with missing values so have been filled with N/A which we need to remove. These missing values are simply from it containing only text in the raw dataset.

```{r cleaning the data, warning=FALSE, message = FALSE}

# First, we need to remove all of the N/A data in the dataset
filtered_data <- data %>% drop_na() 

# To check this has removed all N/A values we use the following function
if (sum(is.na(filtered_data)) == 0) {
  print("No missing values")
} else {
  print(paste("Count of total missing values:", sum(is.na(filtered_data))))
}

# Now we want to remove the percentages data
# The percentage data begins on row 28 in the dataset so we only want to include rows 1 to 27
filtered_data <- filtered_data[1:27, ]

# View the filtered date
head(filtered_data)

# View column names
names(filtered_data)

# Rename a couple of the column names to remove the 'note...' as this is only relevant when viewing the data in Excel
filtered_data <- filtered_data %>% rename("Age of Suspect" = "Age of suspect [note 65]", "Apr 2020 to Mar 2021" = "Apr 2020 to Mar 2021 [note 66]") 

# View column names again to see if the changes have been made
names(filtered_data)
# put in codebook the notes
# separate document which references everything 
```
We remove the N/A values before specifying which rows to keep as the first row of the 'data' dataset only has N/A values, so if we removed the percentages data before we would delete the total values row of the data we're interested in. So to aid visualisation we clean the data by removing N/A values first, and then specify the rows of data we want to analyse and visualise.

# Data Preparation
As the dataset currently has a lot of columns, to be able to visualise it we need to transform it from a wide format to a long format. This is done using the 'pivot_longer' function.

```{r}
data_long <- pivot_longer(data = filtered_data, 
                          names_to = "Year",
                          values_to = "Frequency",
                          cols = c("Apr 2012 to Mar 2013", "Apr 2013 to Mar 2014", "Apr 2014 to Mar 2015", "Apr 2015 to Mar 2016", "Apr 2016 to Mar 2017", "Apr 2017 to Mar 2018", "Apr 2018 to Mar 2019", "Apr 2019 to Mar 2020", "Apr 2020 to Mar 2021", "Apr 2021 to Mar 2022", "Apr 2022 to Mar 2023"))

head(data_long)
```

```{r Make Year column better to visualise}
# Create a string which has just th eyear range rather than month and year so it is easier to visualise on a graph


```

```{r Create a final datafram for the graph}
# Create a new dataframe to use in the visualisation
# Remove the rows which display both male and female so we can compare just male and female
# And include all ages 
sex_data <- data_long %>% filter(`Sex of Suspect` %in% c("Male", "Female") &
                                  `Age of Suspect` == "Total [note 66]")

sex_data <- sex_data %>%
  mutate(
    simple_years = paste(str_sub(Year, 5, 8), str_sub(Year, 16, 20), sep = "-") # Directly create simple_years
  )

sex_data$'Sex of Suspect' <- factor(sex_data$'Sex of Suspect', levels = c("Female", "Male"))
levels(sex_data$'Sex of Suspect')

head(sex_data)
```

# Potential Graph Functions

```{r}

# My barplot:
sex_data$Frequency <- as.numeric(sex_data$Frequency)
p <- ggplot(sex_data, aes(x=simple_years, y=Frequency, fill = `Sex of Suspect`)) +
  geom_bar(stat = "identity", colour='black', width = 0.75) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 550)) +
  scale_x_discrete() + 
  scale_fill_manual(values = c("Male" = "#7A67EE", "Female" = "#FF6A6A")) +
  ggtitle("Number of Homicide Suspects Per Year") + 
  theme(
    panel.background = element_rect(fill = "#BFD5E3", colour = "#6D9EC1", linetype = "solid"),
    axis.text.x = element_text(angle = 45, hjust = 1, family = "sans"),
    axis.text.y = element_text(family = "sans"),  # Set font for y-axis labels
    axis.title.x = element_text(size = 14, face = "plain", family = "sans"),  # Font for x-axis title
    axis.title.y = element_text(size = 14, face = "plain", family = "sans"),  # Font for y-axis title
    legend.title = element_text(size = 14, face = "bold", family = "sans"),  # Font for legend title
    legend.background = element_rect(fill = "#BFD5E3",linetype = 1, colour = "black", linewidth = 1),
    legend.text = element_text(size = 12, family = "sans"),  # Font for legend labels
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 2)
  ) +

  labs(
    x = "Year",        # Label for x-axis
    y = "Number of Suspects",   # Label for y-axis
    fill = "Sex of Suspect")  # Label for the legend
    title = "Number of Homicide Suspects Per Year"

print(p)
```
```{r}
# My barplot:
sex_data$Frequency <- as.numeric(sex_data$Frequency)
p <- ggplot(sex_data, aes(x=simple_years, y=Frequency, fill = `Sex of Suspect`)) +
  geom_bar(stat = "identity", position = "dodge", colour='black', width = 0.75) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 550)) +
  scale_x_discrete() + 
  scale_fill_manual(values = c("Male" = "#7A67EE", "Female" = "#FF6A6A")) +
  ggtitle("Number of Homicide Suspects Per Year") + 
  theme(
    panel.background = element_rect(fill = "#BFD5E3", colour = "#6D9EC1", linetype = "solid"),
    axis.text.x = element_text(angle = 45, hjust = 1, family = "sans"),
    axis.text.y = element_text(family = "sans"),  # Set font for y-axis labels
    axis.title.x = element_text(size = 14, face = "plain", family = "sans"),  # Font for x-axis title
    axis.title.y = element_text(size = 14, face = "plain", family = "sans"),  # Font for y-axis title
    legend.title = element_text(size = 14, face = "bold", family = "sans"),  # Font for legend title
    legend.background = element_rect(fill = "#BFD5E3",linetype = 1, colour = "black", linewidth = 1),
    legend.text = element_text(size = 12, family = "sans"),  # Font for legend labels
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 2)
  ) +

  labs(
    x = "Year",        # Label for x-axis
    y = "Number of Suspects",   # Label for y-axis
    fill = "Sex of Suspect")  # Label for the legend
    title = "Number of Homicide Suspects Per Year"

print(p)
```

# Visualising the Data
```{r}
# Need to group the data together
final_data <- data_long %>%
  group_by(`Sex of Suspect`, `Age of Suspect`)

final_data <- data_long %>%
  group_by(`Sex of Suspect`, `Age of Suspect`) %>%
  summarize(Number = sum(Values, na.rm = TRUE))

ggplot() +
     geom_line(data = data_long, aes(x = "Year", y = "Age of Suspect", color = "Sex of Suspect")) +
     geom_line(data = data_long, aes(x = "Values", y = "Age of Suspect", color = "Sex of Suspect")) +
     labs(title = "Graph of Number of Suspects Each Year by Sex and Age",
          x = "Year",
          y = "Age of Suspect",
          color = "Sex of Suspect") +
     theme_minimal()

ggplot(data, aes(x = "Year", color = "Sex of Suspect", group = "Sex of Suspect")) +
  geom_line(aes(y = "Values"), linewidth = 1) +
  geom_point(aes(y = "Values"), linewidth = 2) +
  geom_line(aes(y = "Age of Suspect"), linetype = "dashed", linewidth = 1) + 
  labs(
    title = "Number of Suspects and Age Trends by Year and Gender",
    x = "Year",
    y = "Number of Suspects",
    color = "Sex of Suspect"
  ) +
  theme_minimal()

ggplot(data, aes(x = "Year", fill = "Values")) +
  geom_histogram(binwidth = 0.5, position = "dodge", color = "black") +
  facet_wrap(~ "Sex of Suspect")
```


# References
Office for National Statistics. (2024, February 08). *Homicide in England and Wales: year ending March 2023.* https://www.ons.gov.uk/peoplepopulationandcommunity/crimeandjustice/articles/homicideinenglandandwales/yearendingmarch2023 
